Type theories as quotient inductive-recursive types

### 2025-02-21 tex/main.tex
Fix "overfull"s

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{float}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}
\lstset{
    breaklines=true,
    breakatwhitespace=true,
    literate=%
      {≡}{$\equiv$}1
      {Γ}{$\Gamma$}1
      {Δ}{$\Delta$}1 
      {→}{$\to$}1
      {≅}{$\cong$}1
}

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given by the QIT below.
%\begin{figure}[b]
\begin{minipage}{.4\textwidth}
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
\end{alignat*}
\end{minipage}
\begin{minipage}{.58\textwidth}
\begin{alignat*}{3}
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}
\end{minipage}
%\caption{The quotient inductive type of join lists}
%\label{fig:QIT-join-list}
%\end{figure}
\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

@@

The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the number of terms in $\sigma$, giving rise to two $\eta$-laws --- one for $\Delta$ being empty, and one for $\Delta$ being non-empty:
\[
  \emptyctx\eta : \implicit{\Gamma, A, \sigma}\; \sigma        =^{\Sub\,\Gamma\,\emptyctx} \; \emptysub
  \qquad\qquad
  \pi\eta       : \implicit{\Gamma, \Delta, \sigma}\; \sigma   =^{\Sub\,\Gamma\,(\Delta, A)} (\pi_1 \sigma, \pi_2 \sigma).
\]
%
The laws for substitution composition are stated using equality constructors, where, for the first time, we see the use of $\transp$ for both sides of the equation to have the same type.
\begin{align*}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} {\sigma ; \idS_{\Delta}} =^{\Sub\,\Gamma\,\Delta} {\sigma}
                                                                                    \qquad\qquad
  \mathsf{idl}    : \implicit{\Gamma, \Delta, \sigma} {\idS_{\Gamma} ; \sigma} =^{\Sub\,\Gamma\,\Delta} {\sigma} \\
  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; (\sigma ; \tau) ; \gamma =^{\Sub\,\Gamma\,\Theta} \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; \sigma ; (\tau , t) =^{\Sub\,\Gamma\,(\Theta, A)} \; (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
\end{align*}
%\begin{alignat*}{5}
%  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
%  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
%  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
%  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; &%& \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} \; &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
%\end{alignat*}
%
Composition of substitutions $\blank;\blank$ acts like list concatenation, so
we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), \sub{t}{\sigma}$. However $\sub{t}{\sigma}$ is of type $\Tm\;\Gamma\;(\sub{\sub{A}{\tau}}{\sigma})$, but the extension $(\sigma; \tau), \sub{t}{\sigma}$ requires $\sub{t}{\sigma}$ to have type $\Tm\;\Gamma\;(\sub{A}{\sigma; \tau})$ instead. Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon \sub{\sub{A}{\tau}}{\sigma} = \sub{A}{\sigma;\tau}$ which we can transport along to fix up the equation.

@@

The type of telescopes $\Tel : \Ctx \to \Set$ over a `base' context $\Gamma$ are defined inductive-recursively together with a concatenation $\blank \dplus\blank : (\Gamma : \Ctx) \to \Tel\;\Gamma \to \Ctx$ as follows.

\begin{minipage}{.6\textwidth}
\begin{alignat*}{5}
  %\Tel               & :                   && \Ctx \to \Set                                     \\
  \emptytel          & : \implicit{\Gamma}\; && \Tel\;\Gamma \\
  \blank,\blank      & : \implicit{\Gamma}\; && (\Xi : \Tel\;\Gamma) \to \Ty\;(\Gamma \dplus \Xi) \to \Tel\;\Gamma
\end{alignat*}
\end{minipage}
\begin{minipage}{.4\textwidth}
\begin{alignat*}{5}
  %\blank & \dplus\blank     && :       (\Gamma : \Ctx) (\Xi : \Tel\;\Gamma) \to \Ctx \\
  \Gamma & \dplus \emptytel && \reduce \Gamma \\
  \Gamma & \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip\noindent
Equipped with telescopes, we can now generalise single lifting with type \eqref{eq:type-of-lifting} to multiple lifting with type
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus \subTel{\Xi}{\sigma})\;(\Delta \dplus \Xi)
\]
which asks for telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$ defined mutually:

\begin{minipage}{.6\textwidth}
\begin{alignat*}{5}
  \subTel{\emptytel}{\sigma} & {} \reduce \emptytel \\
  \subTel{(\Xi, A)}{\sigma}  & {} \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi}
\end{alignat*}
\end{minipage}
\begin{minipage}{.4\textwidth}
\begin{alignat*}{5}
  & \sigma \upuparrows \emptytel && \reduce \sigma \\
  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
\end{alignat*}
\end{minipage}
\vskip\belowdisplayskip

To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows \subTel{\Xi}{\tau}); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus \subTel{\Xi}{\idS})\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
%
Instead of resorting to \transp, we can also turn telescope substitution $\subTel{\blank}{\blank}$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:

\begin{minipage}{.5\textwidth}
\begin{alignat*}{3}
  \subTel{\Xi}{\idS}          & \reduce \Xi \\
  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{\subTel{\Xi}{\tau}}{\sigma}
\end{alignat*}
\end{minipage}
\begin{minipage}{.5\textwidth}
\begin{alignat*}{5}
  & \idS          && \upuparrows \Xi && \reduce \idS \\
  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
\end{alignat*}
\end{minipage}\vskip\belowdisplayskip\noindent
%whose confluence property is easy to see.

@@

We can add extensional identity types, a hierarchy of Coquand universes~\cite{Coquand2018}, and a unit type with an eta law, all without any additional difficulty. In particular, all these additions can be done without introducing any transports in the syntax. However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
We introduce the type $\bool : \implicit{\Gamma}\;\Ty\;\Gamma$ itself, its constructors $\btrue, \bfalse : \implicit{\Gamma}\;\Tm\;\Gamma\;\bool$ and its elimination principle:
%
\begin{alignat*}{3}
  \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}}) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}}) \\
  &&& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad \to (b : \Tm\;\Gamma\;\bool) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{b}})
\end{alignat*}
%
where $\instSub{t} \defeq (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ is the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place. We can add the expected computation rules 
%
%\begin{alignat*}{5}
%  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}})} {} && f \\
%  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}})} {} && t
%\end{alignat*}
%
as constructors, e.g.\ $ \bool\beta_{\bfalse} : \implicit{\Gamma, P, f, t}\; \belim\,P\,f\,t\,\bfalse = f$, and the structural rule for type substitution as another straightforward clause
$\subTy{\bool}{\sigma} \reduce \bool$,
but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following
%
\begin{alignat*}{5}
  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \sub{(\belim\,P\,f\,t\,b)}{\sigma} \\
           & =^{\Tm\,\Gamma}_{\alert{p(b)}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p(\bfalse)}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p(\btrue)}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
\end{alignat*}
where the proof $p(x) : \subTy{P}{\instSub{\sub{x}{\sigma}} ; \sigma^+} = \subTy{P}{\sigma ; \instSub{x}}$ is needed to line up the types. Even though this is an equation between type substitutions, which we could hope make compute, there is no obvious way to direct the equation, and simply adding it as a reduction leads to confluence problems. Hence, we do not know how to get rid of these transports.

>>>>

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{float}
\usepackage{manfnt}
\usepackage{mathpartir}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}
\lstset{
    breaklines=true,
    breakatwhitespace=true,
    literate=%
      {≡}{$\equiv$}1
      {Γ}{$\Gamma$}1
      {Δ}{$\Delta$}1 
      {→}{$\to$}1
      {≅}{$\cong$}1
}

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020}, where constructors can introduce not only new elements but also new equalities between the introduced elements.
%
%Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type $\JList{A}$ of \emph{join lists} can be given below
%\begin{figure}[b]
\begin{mathpar}
  \Jnil : \implicit{A}\; \JList{A} \and
  \Jsing{\blank} : \implicit{A}\; A \to \JList{A} \and
  \Jconcat{\blank}{\blank} : \implicit{A}\; \JList{A} \to \JList{A} \to \JList{A} \and
  \mathsf{idr} : \implicit{\ell}\; (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \and
  \mathsf{idl} : \implicit{\ell}\; (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \and
  \mathsf{assoc} : \implicit{\ell_0, \ell_1, \ell_2}\; \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{mathpar}
%\begin{minipage}{.6\textwidth}
%\begin{alignat*}{3}
%  \Jnil                    & : \implicit{A}\;      && \JList{A} \\
%  \Jsing{\blank}           & : \implicit{A}\;      && A \to \JList{A} \\
%  \Jconcat{\blank}{\blank} & : \implicit{A}\;      && \JList{A} \to \JList{A} \to \JList{A}
%\end{alignat*}
%\end{minipage}
%\begin{minipage}{.4\textwidth}
%\begin{alignat*}{3}
%\end{alignat*}
%\end{minipage}
%\caption{The quotient inductive type of join lists}
%\label{fig:QIT-join-list}
%\end{figure}
%\vskip\belowdisplayskip\noindent
A quotient inductive-inductive type~\cite{Altenkirch2018} (QIIT) further allows the simultaneous definition of a type $A$ and a type family $B : A \to \Set$ defined inductively.
% --- this is different from an inductive-recursive definition of $A$ and $F : A \to D$, where $F$ is defined recursively.
%; indeed, inductive-inductive definitions only make sense when $D = \Set$.
The metatheory of QIITs has been extensively studied by Altenkirch et al.~\cite{Kaposi2018,Kaposi2019,Kovacs2020}.

@@

The codomain $\Delta$ of a substitution $\sigma:\Sub\;\Gamma\;\Delta$ also tells the number of terms in $\sigma$, giving rise to two $\eta$-laws --- one for $\Delta$ being empty, and one for $\Delta$ being non-empty:
\[
  \emptyctx\eta : \implicit{\Gamma, A, \sigma}\; \sigma        =^{\Sub\,\Gamma\,\emptyctx} \; \emptysub
  \qquad\qquad
  \pi\eta       : \implicit{\Gamma, \Delta, \sigma}\; \sigma   =^{\Sub\,\Gamma\,(\Delta, A)} (\pi_1 \sigma, \pi_2 \sigma).
\]
%
The laws for substitution composition are stated using equality constructors, where, for the first time, we see the use of $\transp$ for both sides of the equation to have the same type.
\begin{align*}
  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma}\; {\sigma ; \idS_{\Delta}} =^{\Sub\,\Gamma\,\Delta} {\sigma}
                                                                                    \qquad\qquad
  \mathsf{idl}    : \implicit{\Gamma, \Delta, \sigma}\; {\idS_{\Gamma} ; \sigma} =^{\Sub\,\Gamma\,\Delta} {\sigma} \\
  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; (\sigma ; \tau) ; \gamma =^{\Sub\,\Gamma\,\Theta} \sigma ; (\tau ; \gamma) \\
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; \sigma ; (\tau , t) =^{\Sub\,\Gamma\,(\Theta, A)} \; (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
\end{align*}
%\begin{alignat*}{5}
%  \mathsf{idr}    & : \implicit{\Gamma, \Delta, \sigma} && {\sigma ; \idS_{\Delta}} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
%  \mathsf{idl}    & : \implicit{\Gamma, \Delta, \sigma} && {\idS_{\Gamma} ; \sigma} && =^{\Sub\,\Gamma\,\Delta} && {\sigma} \\
%  \text{;-}\mathsf{assoc} & : \implicit{\Gamma, \Delta, \Theta, \Xi, \sigma, \tau, \gamma}\; && (\sigma ; \tau) ; \gamma && =^{\Sub\,\Gamma\,\Theta} &&  \sigma ; (\tau ; \gamma) \\
%  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} \; &%& \sigma ; (\tau , t)      && =^{\Sub\,\Gamma\,(\Theta, A)} \; &&  (\sigma ; \tau) , \alert{\transfib{\Tm\,\Gamma}{[;]_{\Ty}^{-1}}{\color{black}\sub{t}{\sigma}}}
%\end{alignat*}
%
Composition of substitutions $\blank;\blank$ acts like list concatenation, so
we would expect $\sigma; (\tau, t)$ to be the same thing as $(\sigma; \tau), \sub{t}{\sigma}$. However $\sub{t}{\sigma}$ is of type $\Tm\;\Gamma\;(\sub{\sub{A}{\tau}}{\sigma})$, but the extension $(\sigma; \tau), \sub{t}{\sigma}$ requires $\sub{t}{\sigma}$ to have type $\Tm\;\Gamma\;(\sub{A}{\sigma; \tau})$ instead. Of course, we have an equality constructor $[;]_{\Ty}^{-1}\colon \sub{\sub{A}{\tau}}{\sigma} = \sub{A}{\sigma;\tau}$ which we can transport along to fix up the equation.

@@

The type of telescopes $\Tel : \Ctx \to \Set$ over a `base' context $\Gamma$ are defined inductive-recursively together with a concatenation $\blank \dplus\blank : (\Gamma : \Ctx) \to \Tel\;\Gamma \to \Ctx$ as follows.
\begin{alignat*}{9}
  \emptytel          & : \implicit{\Gamma}\; && \Tel\;\Gamma \qquad 
                     && \Gamma && \dplus \emptytel && \reduce \Gamma \\
  \blank,\blank      & : \implicit{\Gamma}\; && (\Xi : \Tel\;\Gamma) \to \Ty\;(\Gamma \dplus \Xi) \to \Tel\;\Gamma\qquad
                     && \Gamma && \dplus (\Xi, A)  && \reduce (\Gamma \dplus \Xi) , A
\end{alignat*}
Equipped with telescopes, we can now generalise single lifting with type \eqref{eq:type-of-lifting} to multiple lifting with type
\[
\blank\upuparrows \blank : (\sigma : \Sub\;\Gamma\;\Delta) \to (\Xi : \Tel\;\Delta)
\to \Sub\;(\Gamma \dplus \subTel{\Xi}{\sigma})\;(\Delta \dplus \Xi)
\]
which asks for telescope substitution $\subTel{\blank}{\blank} :  \Tel\;\Delta \to \Sub\;\Gamma\;\Delta \to \Tel\;\Gamma$ defined mutually:
\begin{alignat*}{9}
  \subTel{\emptytel}{\sigma} & {} \reduce \emptytel \qquad
                             && \sigma \upuparrows \emptytel && \reduce \sigma \\
  \subTel{(\Xi, A)}{\sigma}  & {} \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi} \qquad\qquad
                             && \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
\end{alignat*}
%\begin{minipage}{.6\textwidth}
%\begin{alignat*}{5}
%  \subTel{\emptytel}{\sigma} & {} \reduce \emptytel \\
%  \subTel{(\Xi, A)}{\sigma}  & {} \reduce \subTel{\Xi}{\sigma}, \subTy{A}{\sigma \upuparrows \Xi}
%\end{alignat*}
%\end{minipage}
%\begin{minipage}{.4\textwidth}
%\begin{alignat*}{5}
%  & \sigma \upuparrows \emptytel && \reduce \sigma \\
%  & \sigma \upuparrows (\Xi, A)  && \reduce (\sigma \upuparrows \Xi) \uparrow A
%\end{alignat*}
%\end{minipage}
%\vskip\belowdisplayskip

To prove the coherence property for the multiple lifting, we would need the properties $\idS_\Gamma \upuparrows \Xi = \idS_{\Gamma \dplus  \Xi}$ and $(\sigma; \tau) \upuparrows \Xi = (\sigma\upuparrows \subTel{\Xi}{\tau}); (\tau \upuparrows \Xi)$, similar to \cref{lem:lifting}, but these identities cannot be typed: the type of $\idS_{\Gamma} \upuparrows \Xi$ is $\Sub\;(\Gamma \dplus \subTel{\Xi}{\idS})\;(\Gamma \dplus \Xi)$ while the type of $\idS$ on the RHS is $\Sub\;(\Gamma \dplus \Xi)\;(\Gamma \dplus \Xi)$; similarly for the substitution composition.
%
Instead of resorting to \transp, we can also turn telescope substitution $\subTel{\blank}{\blank}$ and the multiple lifting into definitions by overlapping patterns by adding the following clauses:
\begin{alignat*}{9}
  \subTel{\Xi}{\idS}          & \reduce \Xi\qquad\qquad
                              && \idS          && \upuparrows \Xi && \reduce \idS \\
  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{\subTel{\Xi}{\tau}}{\sigma} \qquad\qquad
                              && (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
\end{alignat*}
%\begin{minipage}{.5\textwidth}
%\begin{alignat*}{3}
%  \subTel{\Xi}{\idS}          & \reduce \Xi \\
%  \subTel{\Xi}{\sigma ; \tau} & \reduce \subTel{\subTel{\Xi}{\tau}}{\sigma}
%\end{alignat*}
%\end{minipage}
%\begin{minipage}{.5\textwidth}
%\begin{alignat*}{5}
%  & \idS          && \upuparrows \Xi && \reduce \idS \\
%  & (\sigma;\tau) && \upuparrows \Xi && \reduce (\sigma \upuparrows \subTel{\Xi}{\tau}) ; (\tau \upuparrows \Xi)
%\end{alignat*}
%\end{minipage}\vskip\belowdisplayskip\noindent
%whose confluence property is easy to see.

@@

We can add extensional identity types, a hierarchy of Coquand universes~\cite{Coquand2018}, and a unit type with an eta law, all without any additional difficulty. In particular, all these additions can be done without introducing any transports in the syntax. However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
We introduce the type $\bool : \implicit{\Gamma}\;\Ty\;\Gamma$ itself, its constructors $\btrue, \bfalse : \implicit{\Gamma}\;\Tm\;\Gamma\;\bool$ and its elimination principle:
%
\begin{alignat*}{3}
  \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}}) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}}) \\
  &&& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad \to (b : \Tm\;\Gamma\;\bool) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{b}})
\end{alignat*}
%
where $\instSub{t} \defeq (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ is the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place. We can add the expected computation rules 
%
%\begin{alignat*}{5}
%  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}})} {} && f \\
%  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}})} {} && t
%\end{alignat*}
%
as constructors, e.g.\ $ \bool\beta_{\bfalse} : \implicit{\Gamma, P, f, t}\; \belim\,P\,f\,t\,\bfalse = f$, and the structural rule for type substitution as another straightforward clause
$\subTy{\bool}{\sigma} \reduce \bool$,
but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following
%
\begin{alignat*}{5}
  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \sub{(\belim\,P\,f\,t\,b)}{\sigma} \\
           & =^{\Tm\,\Gamma}_{\alert{p\,b}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\bfalse}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\btrue}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
\end{alignat*}
where the proof $p\,x : \subTy{P}{\instSub{\sub{x}{\sigma}} ; \sigma^+} = \subTy{P}{\sigma ; \instSub{x}}$ is needed to line up the types. Even though this is an equation between type substitutions, which we could hope make compute, there is no obvious way to direct the equation, and simply adding it as a reduction leads to confluence problems. Hence, we do not know how to get rid of these transports.
